% Make sure to keep the text within the brackets and the output should turn out correct
\acknowledge{%
  
First and foremost I'd like to thank my advisor, Mikl\'os B\'ona, for his
guidance and encouragement throughout the research process, and for his
patience and understanding during my meandering course through
graduate school. 
I also thank Vince Vatter, whose long discussions, advice, and friendship have
helped make my time here more productive and more enjoyable.  Thanks also
to Michael Albert for his support and suggestions during our collaborations,
and to the remainder of my supervisory committee: Andrew Vince, Meera Sitharam,
and Kevin Keating, each of whom have helped me to become a more well-rounded
researcher. 


This dissertation is a product of the combined support of those around me, each
of whom have left a profound impact both on this work and on my time in
graduate school.  The graduate student community, with its many seminars and
happy hours, has made the last five years more fun than it should have been.  I
am grateful for all of my friends and colleagues, both for their support during
the busy times and for their distractions during the slow.  


My time at the University of Florida has been marked with frequent
diversions ---  organizing seminars and serving on administrative committees
has kept me busy and interested, and I am thankful for all of the coworkers and
friends I've met along the way.  Special thanks to Margaret, Connie, and
the rest of the math department staff for helping me to find more travel
funding than any graduate student deserves.  Finally, I am thankful for my
students, who taught me to never stop looking for a simpler way to present a
problem.

I am grateful for my wonderfully supportive family, who have always always
encouraged me in every endeavor, fostered every interest, and listened to me
long before I had anything to say.  Finally, thank you to my best friend and
favorite travel partner Elizabeth, for her support, editing skills, and
understanding during the last five years, and for pushing me to be better in
every way. 



}
